Controlled Experiments for Word Embeddings

### 2015-12-02 article/main.tex
article changes prior to AS rereading, updating after Schnabel disc

We show that the length of the word vectors generated by the CBOW model
depends more or less linearly on both word frequency and level of noise
in the co-occurrence distribution of the word.  In both cases, the
coefficient of linearity depends upon the word.  If the co-occurrence
distribution is fixed, then word vector length increases with word
frequency.  If, on the other hand, word frequency is held constant, then
word vector length decreases as the level of noise in the co-occurrence
distribution of the word is increased.  We show furthermore that the
direction of the word vectors is independent of word frequency and only
moderately depends upon co-occurrence noise.  When noise is added to the
co-occurrence distribution of a word, the corresponding vector smoothly
interpolates between the original word vector and a small vector
perpendicular to it that represents a word with pure noise in its
co-occurrence distribution.  Surprisingly, the special point in feature
space, obtained by interspersing a pseudoword uniformly at random
throughout the corpus with a frequency sufficiently large to sample all
contexts, is non-zero.

@@

\begin{section}{Related work}\label{related-work}
In recent work on the evaluation of word embeddings, Schnabel et
al.~\cite{schnabelemnlp2015} trained logistic regression models to
predict whether a word was rare or frequent given only the direction of
its word vector.  For various word embedding methods, the prediction
accuracy was measured as a function of the threshold for word rarity.
It was found in the case of word2vec CBOW that word vector direction
could be used to distinguish very rare words from all other words. The
authors claimed further that word vector direction must encode frequency
directly, and not indirectly via semantic information.  Our experimental
finding that word vector direction is independent of word frequency
contradicts this claim in the case of word2vec CBOW.

Our experimental finding that word vector length decreases with co-occurrence
noise is related to earlier work by Vecchi, Baroni, and Zamparelli
\cite{vecchi-baroni-zamparelli2011}, where a relation between vector length and
the ``semantic deviance'' of an adjective-noun composite was studied
empirically.  In that paper, which is also based on word co-occurrence
statistics, the authors study adjective-noun composites.  They built a
vocabulary from the 8k most frequent nouns and 4k most frequent adjectives in a
large general language corpus and added 22k adjective-noun composites.  For
each item in the vocabulary, they recorded the co-occurrences with the top 10k
most frequent content words (nouns, adjectives or verbs), and constructed word
embeddings via singular value decomposition of the co-occurrence matrix
\cite{landauer-dumais1997}.  The authors considered several models for
constructing vectors of unattested adjective-noun composites, the two simplest
being adding and component-wise multiplying the adjective and noun vectors.
They hypothesized that the length of the vectors thus constructed can be used
to distinguish acceptable and semantically deviant adjective-noun composites.
Using a few hundred adjective-noun composites selected by humans for
evaluation, they found that deviant composites have a shorter vector than
acceptable ones, in accordance with their expectation.  In contrast to their
work, our approach does not require human annotation.

@@

Consider first the words other than \word{the} and \word{VOID}.  These
words typically occur in a limited number of distinct contexts, and so a
limited number of samples is required to capture their co-occurrence
distributions.  We therefore expect that the vectors representing the
pseudowords derived from any of these words can be properly learned from the
modified corpus, even though their combined frequency equals the
frequency of that word in the original corpus.  The red blocks in
Figure~\ref{word-frequency-experiment-heatmap} clearly demonstrate that
the direction of these vectors is constant for each family of pseudowords.
In other words, the direction of these vectors is only determined by the
original word and is independent of how often that word appears,
provided sufficient samples have been processed for the word vector to
settle in its equilibrium direction.

The artificial word \word{VOID}, on the other hand, occurs by
construction in every context.  A much higher number of samples is
therefore required to capture its co-occurrence distribution, and
thereby to learn its word vector.  The same is true, but to a lesser
extent, for the stopword \word{the}.  The difficulty of learning the
vector representing \word{VOID\_i} or \word{THE\_i} increases with $i$,
for the pseudowords occur increasingly less often.  This is consistent with
Figure~\ref{word-frequency-experiment-heatmap}, where it is apparent
that the vectors representing \word{VOID\_i} for small $i$, say $i = 1,
\dots, 5$, roughly share a common direction, while this is no longer
true for higher values of $i$.  The same trend is apparent, although
less extremely so, for the word \word{the}.  The vectors for the pseudowords
\word{THE\_i} share a common direction for values of $i$ as high as
$14$, before they start changing direction.  We conclude that these
deviations are artifacts of poor learning due to an insufficient number
of occurrences of these pseudowords.
Table~\ref{word-frequency-experiment-cosine-similarity}, showing the
most similar words to each pseudoword \word{THE\_i}, supports this further.
In the discussions below, the pseudowords \word{VOID\_i} and \word{THE\_i}
whose vectors have a cosine similarity less than $0.8$ with
\word{VOID\_1} and \word{THE\_1}, respectively, are therefore discarded.

From these results, we conclude that word frequency has no effect on the
direction of a word vector, given a sufficient number of samples to
learn the co-occurrence distribution of the word.

@@

\begin{table}
	\input{cosine-similarity.tex}
	\caption{ Words in the original vocabulary most similar to the
          pseudowords \word{THE\_i}, and their cosine similarity with the
          most frequent such pseudoword, \word{THE\_1}.  It is apparent from
          the nearest neighbour list that the vectors of the
          low-frequency pseudowords have not been adequately trained.  }
	\label{word-frequency-experiment-cosine-similarity}
\end{table}

@@

>>>>

We show that the length of the word vectors generated by the CBOW model
depends more or less linearly on both word frequency and level of noise
in the co-occurrence distribution of the word.  In both cases, the
coefficient of linearity depends upon the word.  If the co-occurrence
distribution is fixed, then word vector length increases with word
frequency.  If, on the other hand, word frequency is held constant, then
word vector length decreases as the level of noise in the co-occurrence
distribution of the word is increased.  In addition, we show that the direction
of a word vector varies smoothly with word frequency and the level of
co-occurrence noise.  When noise is added to the
co-occurrence distribution of a word, the corresponding vector smoothly
interpolates between the original word vector and a small vector
perpendicular to it that represents a word with pure noise in its
co-occurrence distribution.  Surprisingly, the special point in feature
space, obtained by interspersing a pseudoword uniformly at random
throughout the corpus with a frequency sufficiently large to sample all
contexts, is non-zero.

@@

\begin{section}{Related work}\label{related-work}
Our experimental finding that word vector length decreases with co-occurrence
noise is related to earlier work by Vecchi, Baroni, and Zamparelli
\cite{vecchi-baroni-zamparelli2011}, where a relation between vector length and
the ``semantic deviance'' of an adjective-noun composite was studied
empirically.  In that paper, which is also based on word co-occurrence
statistics, the authors study adjective-noun composites.  They built a
vocabulary from the 8k most frequent nouns and 4k most frequent adjectives in a
large general language corpus and added 22k adjective-noun composites.  For
each item in the vocabulary, they recorded the co-occurrences with the top 10k
most frequent content words (nouns, adjectives or verbs), and constructed word
embeddings via singular value decomposition of the co-occurrence matrix
\cite{landauer-dumais1997}.  The authors considered several models for
constructing vectors of unattested adjective-noun composites, the two simplest
being adding and component-wise multiplying the adjective and noun vectors.
They hypothesized that the length of the vectors thus constructed can be used
to distinguish acceptable and semantically deviant adjective-noun composites.
Using a few hundred adjective-noun composites selected by humans for
evaluation, they found that deviant composites have a shorter vector than
acceptable ones, in accordance with their expectation.  In contrast to their
work, our approach does not require human annotation.

@@

Consider first the word vectors for the pseudowords associated to the word
\word{the}.  Notice that the cosine similarity of the word vectors for
\word{THE\_1} and \word{THE\_i} decreases monotonically with $i$, while the
cosine similarity of the word vectors for \word{THE\_i} and \word{THE\_18}
increases monotonically with $i$.  Indeed the direction of
the word vector \word{THE\_i} changes systematically, interpolating between the
directions of the word vectors of the highest-frequency pseudoword
\word{THE\_1} and the lowest-frequency pseudoword \word{THE\_18}.  The same
trend is apparent (though over shorter frequency ranges) for all the families of pseudowords
other than that for \word{VOID}.

Consider now the word vectors for pseudowords derived from the meaningless word
\word{VOID}.  The word vectors for \word{VOID\_7}, \dots, \word{VOID\_13}
are approximately orthogonal to one another, just as would be expected from
randomly drawn vectors in a high dimensional space.
As the pseudoword \word{VOID} occurs by
construction in every context, a much higher number of samples is
required to capture its co-occurrence distribution, and
thereby to learn its word vector (the same is true, but to a lesser
extent, for the stopword \word{the}).
We conclude that the vectors corresponding to the lower frequency
pseudowords \word{VOID\_7}, \dots, \word{VOID\_13} have not been trained on a
sufficient number of samples to establish their proper direction.
These vectors are excluded from further analysis.
The vectors for \word{VOID\_1}, \dots, \word{VOID\_6}, on the other hand,
exhibit the smooth change in word vector direction with word frequency described in
the previous paragraph.

In recent work on the evaluation of word embeddings, Schnabel et
al.~\cite{schnabelemnlp2015} trained logistic regression models to
predict whether a word was rare or frequent given only the direction of
its word vector.  For various word embedding methods, the prediction
accuracy was measured as a function of the threshold for word rarity.
It was found in the case of word2vec CBOW that word vector direction
could be used to distinguish very rare words from all other words.
Figure~\ref{word-frequency-experiment-heatmap} is consistent with this finding,
as it is apparent that word vector direction does change gradually with
frequency.  Schnabel et al.\ claim further that word vector direction must
encode word frequency directly, and not indirectly via semantic information.
Figure~\ref{word-frequency-experiment-heatmap}, considered for any particular
experiment word in isolation (e.g. \word{SQUAD}), demonstrates that the variance of word
vector direction with word frequency is indeed independent
of co-occurrence (semantic) information, and thereby provides further evidence
for this claim.

@@

@@

\begin{section}{Acknowledgments}
The authors thank Tobias Schnabel for helpful discussions.
\end{section}
